\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-18 paper.qmd
Add tour implemenation for kang data sets with necessary parameters (need to record the video)
Added video links for PBMC3k and Five Guassian clusters

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/uy9SkAo6gAE>), (b) Model in the 2D space with UMAP (<https://youtu.be/0MJDhHrh_Ug>), (c) Model in the 2D space with PHATE (<https://youtu.be/HbVv0uy0QWk>), (d) Model in the 2D space with TriMAP (<https://youtu.be/2OGwipAjzc8>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/pkokI8d-cBk>).
#| out-width: 100%

@@

### Single-cell RNA-seq data of human

@@

## Import data
#df_2 <- read_rds("data/pbmc/pbmc.rds")
training_data_1 <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_pca_50.rds")
training_data_1 <- training_data_1[, 1:15]
#test_1 <- read_rds("data/pbmc/pbmc_test.rds")

plot_list2_pbmc <- UMAP_pbmc |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) #+
  # guides(color = guide_legend(title = "Cell type"),
  #        color = guide_legend(nrow = 3)) #+
  # annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-nldervisPBMCUMAP

plot_list2_pbmc

<!--
```{r}
#| warning: false
#| echo: false

@@

@@

>>>>

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/x5VPB-wOLm4>), (b) Model in the 2D space with UMAP (<https://youtu.be/Cm1dW6iLyHQ>), (c) Model in the 2D space with PHATE (<https://youtu.be/HB0Y3ilz6sI>), (d) Model in the 2D space with TriMAP (<https://youtu.be/iSVkXIrOrSA>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/O2QBctsc4uM>).
#| out-width: 100%

@@

### Human peripheral blood mononclear cells (PBMCs) data

- Intro to the data set (This data set contains 13,714 features across 2,700 samples within a single assay. The active assay is RNA, with 13,714 features representing different gene expressions.)

- 2622 cells

- Evaluate Festem on DEG detection (Festem enabled identification of often-missed fine cell types)

- Using the Festem-selected genes, identified 10 clusters in the PBMCsk data and annotated them based on the expression of canonical markers

- The 10 clusters included immune cells such as naive CD4 T cells, memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes, Natural Killer (NK) cells, B cells and Dendritic Cells (DC).

- In addition to these common cell types, Festem identified a fine cell type, CD27− CD4+ memory T cells, which were often missed by other methods.

- The CD27. CD4+ memory T cells identified by Festem expressed common marker genes (IL7R
and S100A4) of memory T cells, but did not express CD27. These cells also had downregulated
expression of SELL, CCR7, MAL and LEF1, and upregulated expression of CCL5 (Fig. 4B) and
thus were the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+ memory T
cells were known to be at a more differentiated state and have stronger antigen-recall responses
than their CD27+ counterparts.

## Import data
#df_2 <- read_rds("data/pbmc/pbmc.rds")
training_data_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:15] |>
  mutate(ID = 1:NROW(training_data_pbmc))
#test_1 <- read_rds("data/pbmc/pbmc_test.rds")

plot_list2_pbmc <- UMAP_pbmc |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) #+
  # guides(color = guide_legend(title = "Cell type"),
  #        color = guide_legend(nrow = 3)) #+
  # annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false
num_bins_pbmc <- calculate_effective_x_bins(.data = UMAP_pbmc, x = UMAP1,
                                              cell_area = 1) ##23

shape_value_pbmc <- calculate_effective_shape_value(.data = UMAP_pbmc,
                                                      x = UMAP1, y = UMAP2) ## 0.8772751

## To extract bin centroids
hexbin_data_object_pbmc <- extract_hexbin_centroids(nldr_df = UMAP_pbmc,
                                                      num_bins = num_bins_pbmc,
                                                      shape_val = shape_value_pbmc, x = UMAP1, y = UMAP2)

df_bin_centroids_pbmc <- hexbin_data_object_pbmc$hexdf_data

UMAP_pbmc_with_hb_id <- UMAP_pbmc |>
  dplyr::mutate(hb_id = hexbin_data_object_pbmc$hb_data@cID)

## To generate a data set with high-D and 2D training data
df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), UMAP_pbmc_with_hb_id)

## Averaged on high-D
df_bin_pbmc <- avg_highD_data(.data = df_all_pbmc, column_start_text = "PC")

## Triangulate bin centroids
tr1_object_pbmc <- triangulate_bin_centroids(df_bin_centroids_pbmc, x, y)
tr_from_to_df_pbmc <- generate_edge_info(triangular_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2D_dist(.data = tr_from_to_df_pbmc)

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)

trimesh_removed_pbmc_umap <- remove_long_edges(.data = distance_pbmc, benchmark_value = benchmark_pbmc,
                                                 triangular_object = tr1_object_pbmc, distance_col = distance)

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(size = 2, colour = "#33a02c") +
  #ggtitle("(b)") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/Ia3aL3L-fCk>)"
#| label: fig-nldervisPBMCUMAP
 #| out-width: 100%

plot_list2_pbmc + trimesh_removed_pbmc_umap + 
plot_layout(ncol=2)

<!--
### Single-cell RNA-seq data of human

@@

@@

Human retina: https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-7316

@@

For the Kang lupus dataset (29), we obtained 16 cell types based on clustering of the Festem-
selected genes (Fig. 4E and Fig. S10, S11), including 2 fine cell types that were often missed by
other methods. The 2 cell types were IFNhi CD14+ monocytes with high expression of type I
interferon-related (IFN) genes, and HSP+ CD4 T cells with up-regulated heat-
shock-protein-related (HSP) genes. The HSP+ CD4 T cells might result from
cryopreservation and thawing of T cells (37). Other than devianceFS, other available methods
could not identify these HSP+ CD4 T cells (Fig. S11).

### 2024-01-18 paper.tex
Added video links for PBMC3k and Five Guassian clusters

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\caption{\label{fig-modelfiveGau}Is there a best model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model in the 2D space with tSNE
(\url{https://youtu.be/uy9SkAo6gAE}), (b) Model in the 2D space with
UMAP (\url{https://youtu.be/0MJDhHrh_Ug}), (c) Model in the 2D space
with PHATE (\url{https://youtu.be/HbVv0uy0QWk}), (d) Model in the 2D
space with TriMAP (\url{https://youtu.be/2OGwipAjzc8}), and (e) Model in
the 2D space with PaCMAP (\url{https://youtu.be/pkokI8d-cBk}).}

@@

\hypertarget{single-cell-rna-seq-data-of-human}{%
\subsection{Single-cell RNA-seq data of
human}\label{single-cell-rna-seq-data-of-human}}

https://github.com/XiDsLab/Festem\_paper/tree/main

\citet{Chen2023}

https://satijalab.org/seurat/articles/pbmc3k\_tutorial.html
https://www.cell.com/fulltext/S0092-8674(15)00549-8

In the field of single-cell studies, a common analysis task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories (e.g., example 1). In
clustering workflows, the main objective is to verify the existence of
clusters and subsequently identify them as specific cell types by
examining the expression of ``known'' marker genes. In this context, a
``faithful'' embedding should ideally preserve the topology of the data,
ensuring that cells corresponding to the same cell type are situated
close to the high-dimensional space.

To begin our analysis, we installed the Peripheral Blood Mononuclear
Cells (pbmc) data set obtained from 10x Genomics using the
\texttt{SeuratData} R package \citep{Rahul2019}, which facilitates the
distribution of data sets in the form of Seurat objects
\citep{Yuhan2021}. This data set contains 13,714 features across 2,700
samples within a single assay. The active assay is RNA, with 13,714
features representing different gene expressions. After loading the data
set, we obtained the principal components (PCs) and assessed the
variance explained by each PC. Based on this evaluation, we selected
seven PCs, representing approximately 50\% of the variance in gene
expression, for further analysis.

Next, we employed the UMAP technique with default parameter settings. As
illustrated in \textbf{?@fig-pbmc}, the cell types B and Platelet are
well-separated in the UMAP layout. Moreover, CD14+ Mono, FCGR3A+ Mono,
and DC form a distinct cluster, while Naive CD4 T, NK, Memory CD4 T, and
CD8 T are grouped together in another cluster. The values utilized to
construct the smooth low-dimensional manifold are presented in
\textbf{?@tbl-table02}. The linked video, demonstrating the tour with
the model, showcases the generation of a smooth surface for this
application, enabling a comprehensive exploration of the data's
structure and relationships (see \textbf{?@fig-pbmc\_sc}).

@@

\caption{\label{fig-nldervisPBMCUMAP}2D layout from UMAP (n\_neighbors =
30, min\_dist = 0.3) applied for the PBMC3k dataset. Is this a best
representation of the original data?}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

\caption{\label{fig-modelfiveGau}Is there a best model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model in the 2D space with tSNE
(\url{https://youtu.be/x5VPB-wOLm4}), (b) Model in the 2D space with
UMAP (\url{https://youtu.be/Cm1dW6iLyHQ}), (c) Model in the 2D space
with PHATE (\url{https://youtu.be/HB0Y3ilz6sI}), (d) Model in the 2D
space with TriMAP (\url{https://youtu.be/iSVkXIrOrSA}), and (e) Model in
the 2D space with PaCMAP (\url{https://youtu.be/O2QBctsc4uM}).}

@@

\hypertarget{human-peripheral-blood-mononclear-cells-pbmcs-data}{%
\subsection{Human peripheral blood mononclear cells (PBMCs)
data}\label{human-peripheral-blood-mononclear-cells-pbmcs-data}}

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\end{itemize}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/Ia3aL3L-fCk})}
